---
title : "Modelling excess covid death in France"
---
# Preliminaries
*import the packages that we need*

```{r, results = "hide", message=FALSE} 
library(tidyverse)
library(lubridate)
library(aweek)
library(gridExtra)
library(ggplot2)
```

## Function to download OWID data
*This function download the OWID data of the specified country*

```{r}
download_owid_data <- function(country_code) {
  if (country_code == "France") {
    owid_url <- "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/jhu/new_deaths.csv"
  } else {
    owid_url <- "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/jhu/new_deaths.csv"
  }
  
  owid_data <- read_csv(owid_url) %>%
    select(date, !!country_code) %>%
    mutate(!!country_code := ifelse(is.na(!!rlang::sym(country_code)), 0, !!rlang::sym(country_code))) %>%
    mutate(cdeath2 = cumsum(!!rlang::sym(country_code))) %>%
    rename(deaths2 = !!country_code)
  
  return(owid_data)
}
```

```{r, results = "hide"}
owid_data_FRA <- download_owid_data("France")
```

## Function to download JHU data
*This function download the JHU data of the specified country*

```{r}
download_jhu_data <- function(country_name) {
  jhu_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
  jhu_data <- read_csv(jhu_url)
  
  jhu_data <- jhu_data %>%
    rename(Country = `Country/Region`)
  
  jhu_data <- jhu_data %>%
    filter(Country == country_name)
  
  jhu_data <- jhu_data %>%
    select(-Lat, -Long)
  
  jhu_data <- jhu_data %>%
    pivot_longer(cols = -(1:2), names_to = "date", values_to = "cdeaths") %>%
    mutate(date = as.Date(date, tryFormats = c("%m/%d/%Y"))) %>%
    mutate(date = date + dyears(2000) - ddays(15)) %>%
    group_by(date) %>%
    summarise(cdeaths = sum(cdeaths)) %>%
    ungroup() %>%
    mutate(death = cdeaths - lag(cdeaths))
  
  return(jhu_data)
}
```

```{r, results = "hide"}
jhu_data_France <- download_jhu_data("France")
```

## Daily deaths for JHU and OWID
*This code compute the daily deaths for OWID and JHU data frames and plot a graphic*

```{r, warning=FALSE}
daily_deaths_JHU <- jhu_data_France$death

daily_deaths_OWID <- owid_data_FRA$deaths2

comparison_data <- data.frame(
  Date = jhu_data_France$date,
  JHU_Daily_Deaths = daily_deaths_JHU,
  OWID_Daily_Deaths = daily_deaths_OWID
)

ggplot(comparison_data, aes(x = Date)) +
  geom_line(aes(y = JHU_Daily_Deaths, color = "JHU"), linetype = "solid") +
  geom_line(aes(y = OWID_Daily_Deaths, color = "OWID"), linetype = "dashed") +
  labs(title = "Comparaison des Décès Quotidiens (France)",
       y = "Nombre de Décès Quotidiens",
       color = "Source") +
  scale_color_manual(values = c("JHU" = "blue", "OWID" = "red")) +
  theme_minimal()
```

## Cumuluative deaths for JHU and OWID data
*This code compute the cumulative deaths for OWID and JHU data frames and plot a graphic*

```{r}
cumulative_deaths_JHU <- jhu_data_France$cdeaths

cumulative_deaths_OWID <- owid_data_FRA$cdeath2

comparison_data_cumulative <- data.frame(
  Date = jhu_data_France$date,
  JHU_Cumulative_Deaths = cumulative_deaths_JHU,
  OWID_Cumulative_Deaths = cumulative_deaths_OWID
)

ggplot(comparison_data_cumulative, aes(x = Date)) +
  geom_line(aes(y = JHU_Cumulative_Deaths, color = "JHU"), linetype = "solid") +
  geom_line(aes(y = OWID_Cumulative_Deaths, color = "OWID"), lty = 2, lwd = 2) +
  labs(title = "Cumulative death in France",
       y = "Death count",
       color = "Source") +
  scale_color_manual(values = c("JHU" = "black", "OWID" = "red")) +
  guides(color = guide_legend(title = NULL))
  

```

## Function to compute the population
*This function compute the population of a country with the OWID covid dataset*

```{r}
get_population_from_owid <- function(country_name) {
  owid_population_url <- "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv"
  owid_population_data <- owid_population_url %>%
    url() %>%
    read_csv()
  
  population <- owid_population_data %>%
    filter(location == country_name) %>%
    select(population)
  
  return(population)
}
```

```{r, results = "hide"}
population_France <- get_population_from_owid("France")
```

## Function to download mortality data
*This function download the mortality data set*

```{r}
download_mortality_data <- function(country_iso3c) {
  world_mort_url <- "https://raw.githubusercontent.com/akarlinsky/world_mortality/main/world_mortality.csv"
  
  mortality_data <- world_mort_url %>%
    url() %>%
    read_csv()
  
  country_mortality_data <- mortality_data %>%
    filter(iso3c == country_iso3c)
  
  return(country_mortality_data)
}
```

```{r, results = "hide"}
mortality_France <- download_mortality_data("FRA")
```

```{r}
mortality_France %>%
    select(-c(iso3c,country_name,time_unit)) %>%
    rename(mortality=deaths, week=time) ->
    mortality_France
```

## Weekly deaths for France
*We compute weekly deaths for France*

```{r, results = "hide", message=FALSE}
owid_data_FRA <- owid_data_FRA %>%
  mutate(date = as.Date(date))

owid_data_weekly <- owid_data_FRA %>%
  group_by(year = lubridate::year(date), week = lubridate::isoweek(date)) %>%
  summarize(deaths = sum(deaths2))
```

```{r}
owid_data_weekly
```

```{r}
owid_data_weekly <- inner_join(owid_data_weekly,
                                 mortality_France, by=c("year", "week") )

owid_data_weekly %>%
    mutate(start_date = get_date(week=week,year=year)) ->
    owid_data_weekly

tail(owid_data_weekly)
```

```{r}
population_France_constante <- unique(population_France$population)

owid_data_weekly <- owid_data_weekly %>%
  mutate(deaths_per_100k = (deaths / population_France_constante) * 100000)

tail(owid_data_weekly)
```

*We have created a new object for the population, since it is constant and makes calculations easier*
*So we can plot two graphs*

```{r, figures-side, fig.show="hold", out.width="75%"}
p1 <- ggplot(owid_data_weekly, aes(x=start_date))+
  geom_line(aes(y = deaths_per_100k), lwd = 0.7)+
  ggtitle("France covid deaths per 100 thousand")+
  labs(x="date", y="death count")

p2 <- ggplot(owid_data_weekly, aes(x=start_date))+
  geom_line(aes(y = mortality), lwd = 0.7) +
  ggtitle("France : weekly mortality (all causes)")+
  labs(x="date", y="death count")

print(p1)
print(p2)
```

## Function to compute excess deaths
*We use an OLS model to estimate excess deaths in France. The model is modf <- mortality ~ year + factor(week)*

```{r}
mortality_France <- mortality_France %>%
  mutate(start_date = as.Date(paste(year, "01-01"), format = "%Y %d-%m") + weeks(week - 1))
```

```{r}
get_excess_deaths <- function(mortality_data, start_date = "2020-03-01") {
  modf <- mortality ~ year + factor(week) 
  
  ts_model <- lm(modf, data = mortality_data, subset = start_date <= "2020-03-01")

  mortality_data_filtered <- mortality_data[mortality_data$start_date >= start_date, ]
  
  predicted_deaths <- predict(ts_model, newdata = mortality_data_filtered)

  excess_deaths <- mortality_data_filtered$mortality - predicted_deaths

  excess_deaths_per_100k <- (excess_deaths / population_France_constante) * 100000

  mortality_data_filtered$excess_deaths <- excess_deaths
  mortality_data_filtered$excess_deaths_per_100k <- excess_deaths_per_100k

  return(mortality_data_filtered)
}
```

```{r}
mortality_France_with_excess <- get_excess_deaths(mortality_France)
```

*So we can plot two graphs to illustrate*

```{r, figures_side, fig.show='hold', out.width="75%"}
p3 <- ggplot() +
  geom_line(data = owid_data_weekly, aes(x = start_date, y = deaths, color = "covid")) +
  geom_line(data = mortality_France_with_excess, aes(x = start_date, y = excess_deaths, color = "excess deaths")) +
  labs(
    title = "France : covid registred v. excess deaths",
    x = "date",
    y = "death count"
  ) +
  scale_color_manual(values = c("covid" = "black", "excess deaths" = "red"))+
  guides(color = guide_legend(title = NULL)) +
  theme(legend.position = "bottom")

p4 <- ggplot() +
  geom_line(data = owid_data_weekly, aes(x = start_date, y = deaths_per_100k, color = "covid")) +
  geom_line(data = mortality_France_with_excess, aes(x = start_date, y = excess_deaths_per_100k, color = "excess deaths")) +
  labs(
    title = "France : covid registred v. excess deaths per 100 thousand",
    x = "date",
    y = "death count"
  ) +
  scale_color_manual(values = c("covid" = "black", "excess deaths" = "red")) + 
  guides(color = guide_legend(title = NULL)) + 
  theme(legend.position = "bottom")

print(p3)
print(p4)
```

# Country-level analysis
*We now turn our attention to Germany, and apply our functions and have graphs and be able to compare France and Germany*

*We use previous functions to get all the data that we need*

```{r}
owid_data_GER <- download_owid_data("Germany")
```

```{r}
mortality_Germany <- download_mortality_data("DEU")
```

```{r}
population_Germany <- get_population_from_owid("Germany")
population_Germany_constante <- unique(population_Germany$population)
```

*Now we do some arrangements in the data*

```{r, results='hide', message=FALSE}
owid_data_GER <- owid_data_GER %>%
  mutate(date = as.Date(date))

owid_data_GER_weekly <- owid_data_GER %>%
  group_by(year = lubridate::year(date), week = lubridate::isoweek(date)) %>%
  summarize(deaths = sum(deaths2))
```

```{r}
owid_data_GER_weekly
```

```{r}
mortality_Germany <- mortality_Germany %>%
  select(-iso3c, -country_name, -time_unit)
```

```{r}
mortality_Germany <- mortality_Germany %>%
  rename(week = time)
```

```{r}
mortality_Germany <- mortality_Germany %>%
  rename(mortality = deaths)
```

```{r}
owid_data_GER_weekly <- owid_data_GER_weekly %>%
  mutate(start_date = as.Date(paste(year, "01-01"), format = "%Y %d-%m") + weeks(week - 1))
```

```{r}
mortality_Germany <- mortality_Germany %>%
  mutate(start_date = as.Date(paste(year, "01-01"), format = "%Y %d-%m") + weeks(week - 1))
```

*So now we can use this functions*

```{r}
mortality_Germany_with_excess <- get_excess_deaths(mortality_Germany)
```

*And we can plot some graphs*

```{r}
p5 <- ggplot() +
  geom_line(data = owid_data_GER_weekly, aes(x = start_date, y = deaths, color = "covid")) + 
  geom_line(data = mortality_Germany_with_excess, aes(x = start_date, y = excess_deaths, color = "excess deaths")) +
  labs(
    title = "Germany : covid registred v. excess deaths",
    x = "date",
    y = "death count" 
  ) +
  scale_color_manual(values = c("covid" = "black", "excess deaths" ="red")) +
  guides(color = guide_legend(title = NULL)) + 
  theme(legend.position = "bottom")
  
print(p5)
```

# Comparison: excess deaths per 100k

```{r}
p6 <- ggplot() +
  geom_line(data = mortality_France_with_excess, aes(x = start_date, y = excess_deaths_per_100k, color = "France")) + 
  geom_line(data = mortality_Germany_with_excess, aes(x = start_date, y = excess_deaths_per_100k, color = "Germany")) +
  labs(
    title = "Excess deaths per 100 thousand : France and Germany",
    x = "date",
    y = "death count" 
  ) +
  scale_color_manual(values = c("France" = "black", "Germany" ="red")) +
  guides(color = guide_legend(title = NULL)) + 
  theme(legend.position = "bottom")
print(p6)
```